---
title: "DAP Project"
author: "Sparsh Verma"
date: "`r Sys.Date()`"
output:
  html_document: default
  pdf_document: default
---

### Loading the dataset.

```{r}
setwd("C:/Projects/DAP/gdp-growth-analysis")
library(readxl)

df <- read_excel("./data/GDPcurrent-USD-countries.xlsx")


head(df)

str(df)

colnames(df)

```

### converting the values in dataset to billion scale.

```{r}

library(dplyr)

# Step 1: Select columns that are purely numeric (years)
years <- names(df)[grepl("^\\d+$", names(df))]

# Step 2 & 3: Convert to numeric, then to billions and round to 3 decimals
df <- df |>
  mutate(across(all_of(years), ~ round(as.numeric(.x) / 1e9, 3)))

# Preview
head(df)

```

### Finding missing values in dataset.

```{r}

sum(is.na(df))
colSums(is.na(df))

```

### Reshaping the dataset to longer format.

```{r}

library(tidyr)

df_long = pivot_longer(
  df,
  cols = '1970' : '2024',
  names_to = "Years",
  values_to = "Value"
)

df_long

head(df_long)
str(df_long)

df_long$Years <- as.numeric(df_long$Years)

```

### handling missing values now

```{r}
colSums(is.na(df_long))

library(dplyr)
library(zoo)

df_clean = df_long |>
  group_by(Country, IndicatorName) |>
  arrange(Years, .by_group = TRUE) |>
  mutate(Value = na.approx(Value, na.rm = FALSE)) |>
  fill(Value, .direction = "downup")

colSums(is.na(df_clean))

```

### Filtering data for India and "GDP growth (annual %)"

```{r}

df_india <- df_clean |>
  filter(Country == "India", IndicatorName == "Gross Domestic Product (GDP)")

df_india <- df_india |>
  mutate(Value = Value / 1e9)

df_india <- df_india |>
  rename(`Value in Billions` = Value)

head(df_india)

```

### Time Series Analysis for India's GDP from 1970 to 2024

```{r}

library(ggplot2)

ggplot(df_india, aes(x = Years, y = `Value in Billions`)) +
  labs(
    title = "GDP Trends of India from 1970 - 2024",
    x = "Years",
    y = "GDP in Billions"
  ) +
  geom_line()

```

### Transforming values to its log values for converting ecponential to linear trend and making modelling easier.

```{r}

df_india <- df_india |>
  mutate(`Log_GDP` = log(`Value in Billions`))

head(df_india)

ggplot(df_india, aes(x = Years, y = Log_GDP)) +
  labs(
    title =  "GDP Trends of India with Log Values of GDP",
    x = "Years",
    y = "Log GDP"
  ) +
  geom_line()

```

### Creating a feature to separate pre 1991 phase and post 1991 phase for better model and trend analysis because we are having different slopes for growth in both phases.

```{r}

df_india <- df_india |>
  mutate(Post_1991 = ifelse(Years >= 1991, "Yes", "No"))

head(df_india)

table(df_india$Post_1991)

ggplot(df_india, aes(x = Years, y = Log_GDP, color = Post_1991, group = 1)) +
  labs(
    title = "GDP Trends of India with Log Values",
    x = "Years",
    y = "Log GDP",
    color = "Post 1991"
  ) +
  geom_line()

```

### Now filtering out the data with GDP indicator only.

```{r}

df_gdp <- df_clean |>
  filter(IndicatorName == "Gross Domestic Product (GDP)")

# Renaming the columns properly

df_gdp <- df_gdp |>
  rename(
    Year = Years,
    GDP = Value
  )


df_gdp

```

### Now converting the final dataset into long format
### All of the Feature Engineering done in Python

```{r}

library(readxl)
library(tidyverse)

df_final <- read_excel("C:/Projects/DAP/gdp-growth-analysis/data/data_gdp_final.xlsx")

colnames(df_final)

## converting into long format

df_final_long <- df_final |>
  
  # Convert wide → long
  pivot_longer(
    cols = matches("^\\d{4}$"),
    names_to = "Year",
    values_to = "GDP"
  ) |>
  
  # Fix types
  mutate(
    Year = as.integer(Year),
    GDP = as.numeric(GDP)
  ) |>
  
  # Sort before growth calculation (CRITICAL)
  arrange(Country, Year) |>
  
  group_by(Country) |>
  mutate(
    Growth = (GDP - lag(GDP)) / lag(GDP) * 100
  ) |>
  
  mutate(
    Growth = ifelse(row_number() == 1, 0, Growth)
  ) |>
  
  # Reorder columns EXACTLY as you want
  select(
    CountryID,
    Country,
    Year,
    GDP,
    Growth,
    CAGR,
    Growth_Volatility,
    GDP_Multiplier
  ) |>
  
  mutate(Growth = replace_na(Growth, 0))
```

### Adding Rank: Ranking of countries based on GDP within each year
Rank = 1 → Largest economy that year
Rank = 2 → Second largest
etc.

```{r}

df_final_long <- df_final_long |>
  group_by(Year) |>
  mutate(
    Rank = rank(-GDP)
  ) |>
  ungroup()

```

### Adding Ranking for other factors as well

```{r}

df_final_long <- df_final_long |>
  group_by(Year) |>
  mutate(
    Rank_Growth = rank(-Growth)
  ) |>
  ungroup() |>
  
  group_by(Country) |>
  mutate(
    Rank_CAGR = rank(-CAGR),
    Rank_Stability = rank(Growth_Volatility)
  ) |>
  ungroup()

```


### Exporting the dataset
install.packages("writexl")   # run once

```{r}
#ensuring the correct data structure for dataset.
str(df_final_long)

library(writexl)
write_xlsx(df_final_long, "gdp_powerbi_ready.xlsx")

```

